---
title: "Algonquin Turtle Life History Project - Painted Turtle Survival Analysis"
author: "Haoluan Chen"
date: "2/13/2021"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(marked)
```

```{r include=FALSE}
data <- read_excel("turtles_annual_data.xls")

## Data Cleaning 
data1 <- data %>% 
  select("notch", "Year", 'age',
         "sex", "date", "Julian date", "site", "gravid", "nest",
         "nestdate", "secnest", "nest2dat", "firstcap", "mass", "cl",
         "pl", "midcl", "midpl", "lclaw", "rclaw", "cw", "ch", "dead") %>% 
  filter(site == "Arowhon Rd." | site == "Sims Creek" | site == "Maiden Lake"
         | site ==  "Road to WHP" | site == "Blanding's Bog" 
         | site == "Wolf Howl Pond" | site == "Wolf Howl Pond E." 
         | site == "Road to Wrose" | site == "West Rose" 
         | site == "March Hare Lake" | site == "Wolf Howl Pond e.") %>% 
  mutate(site = ifelse(site == "Wolf Howl Pond e.","Wolf Howl Pond E.", site ))

data1 <- data1 %>% 
  mutate(mature = ifelse(sex == "Male" & pl > 9, TRUE, FALSE),
         mature = ifelse(sex == "Female" & pl > 12, TRUE, FALSE),
         mass = ifelse(mass == -47, NA, mass),
         firstcapyear = ifelse(firstcap == TRUE, Year,0))
#data1 is cleaned data

## CH 
turtles <- data %>% select(notch, Year) %>% filter(Year >= 1991)
turtles %>% distinct(Year)
history <-  turtles %>% mutate(capture = 1) %>% 
  complete(notch, Year) %>% 
  mutate(capture = !is.na(capture),
         capture = ifelse(capture == T, 1, 0)) %>% 
  group_by(notch) %>% 
  summarise(ch = paste(capture,collapse = ""))


## Data Manupliation and Transformation
datacov <- data1 %>% select(notch, Year, sex, midpl, dead) # select wanted variables
datawithcapyear <- data1 %>% select(notch, firstcapyear) %>% filter(firstcapyear != 0)

data2 <- full_join(datacov, history, by = "notch")
# cleaning and transforming midpl variable
data2 <- data2 %>% filter(!is.na(midpl)) 
midpl <- data2 %>% group_by(notch) %>% spread(Year, midpl)
midpl <- full_join(midpl, datawithcapyear, by = "notch") %>% filter(!is.na(firstcapyear))


# imputation
year = 1991
column = 5
for(c in column:31) {
    for (i in 1:nrow(midpl)){
      if (midpl[i,32] > year){ 
        midpl[i, c] = 0 
      }
      # impute by assumping each year turtles' midpl grow by 0.15 
      else if (is.na(midpl[i,c])){
        if (year > 1991 && midpl[i,32] < year){
        midpl[i,c] = midpl[i, c-1] + 0.15
        }
      }


      # impute by previous observation
      # if(year > 1991 && is.na(midpl[i,c])){
      #   midpl[i,c] = midpl[i, c-1]
      # }
    }
  year = year + 1 

}
midpl

# renaming the columns
midpl <- rename(midpl, c("midpl1" = "1991", "midpl2" = "1992", "midpl3" = "1993", "midpl4" = "1994", "midpl5" = "1995", "midpl6" = "1996", "midpl7" = "1997", "midpl8" = "1998", "midpl9" = "1999", "midpl10" = "2000", "midpl11" = "2001", "midpl12" = "2002", "midpl13" = "2003", "midpl14" = "2004", "midpl15" = "2005", "midpl16" = "2006", "midpl17" = "2007", "midpl18" = "2008", "midpl19" = "2009", "midpl20" = "2010", "midpl21" = "2011", "midpl22" = "2012", "midpl23" = "2013", "midpl24" = "2014", "midpl25" = "2015","midpl26" = "2016", "midpl27" = "2017"))

midpl <- na.omit(midpl)



 # Process data
turtles.proc = process.data(midpl)
turtles.proc

 # Create design data with static and time varying covariates
design.Phi=list(static=c("sex"))
design.p=list(static=c("sex")) 

 design.parameters=list(Phi=design.Phi,p=design.p)
 ddl=make.design.data(turtles.proc,parameters=design.parameters) 
 names(ddl$Phi)
 names(ddl$p)
 
Phi.sfw=list(formula=~sex)
p.ast=list(formula=~sex)
model=crm(turtles.proc,ddl,hessian=TRUE,model.parameters=list(Phi=Phi.sfw,p=p.ast))


newdipper=expand.grid(sex=c("Female","Male", "Unknown"))

pred = predict(model, newdata = newdipper, se= TRUE)

ggplot(pred$Phi, aes(sex, estimate, ymin=lcl, ymax=ucl))+geom_errorbar(width=0.2) + geom_point()
```

# Abstract

# Introduction
** Reptiles are declining globally, the turtles is one of the most threatened species on the planet.**

Algonquin Park is a beautiful place that attracts many tourists to came and enjoyed these wonderful views. At the same time Algonquin park is home to many animal species. Every year starting from 1978 a team of dedicated researchers travel to the Wildlife Research Station in Algonquin Provincial Park, where they live from May to August. The team captured and recaptured the painted turtle then assigned unique notch identification and recorded their size and other statistics. Their goal is to use this long-term dataset to address interesting and timely questions, such as adaptation to northern environments, life-history trade-offs, phenotypic response to climate change, sexual selection, among other topics. However, for our study, we are investigating the relationship between the painted turtles survival rate and its size together with other behaviors. 

In Rest of my report contains methods, which I first explain my data cleaning decision and model selection process in the Methods section. Then, I discuss the model result in the Results section, follow by model implication in the Discussion section. Finally in my conclusion, I recap our finding through our statistical analysis and concludes our result regarding to the painted turtles survival rate in the wild.

# Methods

## Data Cleaning dicision
Based on the information provided from our collaborator, that the sex and size variables are likely to have a impact on the survival rate of the painted turtle. So, I have removed unreliable and Irrelevant variables in the dataset and we left with the following remaining 16 variables.

Variables | Definition
------------- | -------------
notch | Unique individual ID 
Year | Year that measurements were taken
sex | Either UNKNOWN (immature individuals can't be sexed) or Male, or Female)
date | The date of measurement
site | site that the turtle was captured
firstcap | Whether or not this is the first time we have ever caught that turtle in the history of the project
mass | weight of the turtle 
cl | carapace length (maximum length)
pl | Plastron Length (maximum length)
midcl | Carapace length (Measured from between the anterior-most carapacial scutes to between the postior-most carapacial scutes)
midpl | Plastron length (Measured from between the anterior-most plastral scutes to between the postior-most plastral scutes)
lclaw | Length of the third nail of the left claw
rclaw | Length of the third nail of the right claw
cw | The width of the widest portion of the carapace
ch | The height of the turtle, measured from the bottom of the plastron to the top of the carapace.
dead | Whether the turtle was found dead 




## Model Selection

Our dataset was obtained using capture and recapture methods. This method cause a high correlation in observations from the same turtle, which made me realized those common statistical models with independent assumption with independent assumpation did not apply to our dataset. Additionally, using generalized mixed effect model did not coverage because there is only around 20 dead turtle observation in our dataset. However, we discovered Cormack-Jolly-Seber (CJS) model might be the most appropriate model to fit our dataset. The CSJ model is capable of estimating the survival and the capture probability of individuals and in a population and identifying factors that affected the survival rate and capture rate using the capture-recapture dataset. Based on those features of the CJS model is the best model in our case. 

Should I talk about this? --After further investigation, I found that all the size measurements are highly correlation, and putting all of them into a model would cause co-linearity problem. According to our collaborator, the midpl are the most reliable measurement, so I chose to use only midpl in my model to represent the size of the turtle. 

After building several model with different covariates, I found that the best model is actually using only sex as static variable for both survival probability and capture probability.

## Data Manpulation and imputation Process for CJS model -- Do we need this? Size is not in my final model. what do I do.
For CJS model, it is required to have observations for each turtle every year, but we only have observations from the turtles that is capture on a given year. Therefore, I decided to impute the ...

# 3. Result
```{r include=FALSE}
model$results
```

```{r echo=FALSE}
coefficient <- c("Phi.(Intercept)","Phi.sexMale","Phi.sexUnknown","p.(Intercept)","p.sexMale","p.sexUnknown")
Estimate <- c(3.7025814, -0.6943481, -3.4392258, 1.2086133, -0.2454782, -0.5087132)
se <- c(0.09164597,0.15513322 ,0.13280805	,0.03344518,0.06994550,0.17254081	)
df <- data.frame(coefficient,Estimate,se )
df <- df %>% mutate(lcl = Estimate - 2*se, ucl = Estimate + 2*se)
df <- df %>% mutate(Estimate = round(exp(Estimate),3), lcl = round(exp(lcl),3), ucl = round(exp(ucl),3)) %>% select(-se)
knitr::kable(df)
```

From the above table, we see that the odds ratio between the male turtle and female turtle survival rate is 0.499 with a 95% confidence interval of [0.366, 0.681]. This means the odds of survival rate for male turtle are about 50% lower than the odds of survival rate for female. Furthermore, the odds ratio between the Unknown sex turtle (juvenile) and the female turtle survival rate is 0.032 with a 95% confidence interval of [0.025, 0.042]. This means the odds of survival rate for Unknown sex turtles is 96.8% lower than the odds of survival rate for the female turtle.  

Additionally, looking at the capture probability p, we see that male turtle has lower odds than female turtle, and unknown sex turtle has even lower odds ratio. The odds ratio of capture probability between male and female is 0.782 with a 95% confidence interval of [0.680, 0.9]. And the odds ratio of capture probability between unknown sex and female is 0.601 with a 95% confidence interval of [0.426, 0.849]

```{r echo=FALSE,fig.height=4}
newdipper=expand.grid(sex=c("Female","Male", "Unknown"))
pred = predict(model, newdata = newdipper, se= TRUE)
ggplot(pred$Phi, aes(sex, estimate, ymin=lcl, ymax=ucl))+geom_errorbar(width=0.2) + geom_point() + labs(title = "Painted Turtle Predicted Survival Probability") + ylab("Survival Rate")
```

The odds of survival rate for female, male and unknown sex turtle are also reflected on the above plot. We predicted that female turtles have a high survival probability with a small 95% confidence interval. Male turtles have a slightly lower survival rate and slightly wider 95% confidence interval. Lastly, the unknown sex turtle has the lowest survival rate with only around 57%

```{r echo=FALSE, fig.height=4}
ggplot(pred$p, aes(sex, estimate, ymin=lcl, ymax=ucl))+geom_errorbar(width=0.2) + geom_point() + labs(title = "Painted Turtle Predicted Capture Probability") + ylab("Capture Rate")
```

Our model also predicted that the female turtle has the highest capture probability, around 77%. Male turtles have a slightly lower capture probability, around 72.5%. Lastly, the unknown sex turtle has the lowest estimated capture probability of 67%, with a very wide 95% confidence interval.

# 4. Disscussion 

## Result implication
From our CJS model result, we see that the female turtle has the highest survival probability. Male turtles have a slightly lower survival probability. Unsurprisingly, unknown sex turtles have a very low survival probability compared to male and female turtles. This is consistent with our knowledge of painted turtles since the female turtle is generally larger than males, so they are much less vulnerable. And the unknown sex turtle are mostly juvenile, which are very vulnerable. 

We also observe the same trend for the capture probability that the female turtle has the highest capture probability, followed by male turtle then, unknown sex turtle. This means that we only capture about 77% of female turtles, 72.5% of male turtles, and only 67% of unknown sex turtle. This means we are likely to overestimate the survival probability because we do not have the data for the whole population. There are estimated 1/3 of juveniles are missing from our data, because they are likely to be dead and been eaten. 

## Model limitation

Our model has several limitations. First of all, there were many missing values in our dataset, and we have to remove some of the missing values and perform imputation. Removing and imputing the data could have a severe impact on the model result. Secondly, there was some input error in the data, which I chose to remove. This introduces more variation in our model result. Thirdly, some of our assumptions were not satisfied. For example, we cannot ensure that turtle emigrates from the study area is permanent and we are not releasing the turtles right away. Lastly, our model could not take advantage of all the size measurements and other variables in the dataset. Since adding a size measurement as a time-varying covariate leads to a non-significant result, which may because sex and size measurements are highly correlated. And I was not able to add a site random effect to our CJS model.   


# Conclusion

In conclusion, our model shows female turtles have a very high survival rate and capture rate. Male turtles have a slightly lower survival rate and capture compared to females. And the unknown sex (juveniles) turtle has a much lower survival rate probability and capture probability than female and male. However, there is some limitation of the model due to missing value, input error, imputation methods and the assumption that we may need further investigation. 

# Acknowledgments

# References 

# Appendixes 

